k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-05-08 paper.Rmd
more editing

In the following, I present a variation of the $\mathit{SSE}$ that can be used to minimize differences in variances between groups. To motivate the new criterion, we first consider a set of one-dimensional observations. The sample variance of a set of $N$ observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

$$
\mathit{Var(x)} = \frac{1}{N} \sum\limits_{i=1}^{N} (x_i - \overline{\mathbf{x}})^2
$$

When defining a new variable $y = (y_1, \ldots, y_N)$ as the squared deviation of each observation from the mean $\overline{\mathbf{x}}$, i.e. $y_i = (x_i - \overline{\mathbf{x}})^2$, the sample variance of $x$ can be reformulated as the mean of $y$:

@@

By maximizing $\mathit{SSE_{kplus}}$ we seek to obtain a partitioning where means as well as variances are similar between groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of means and variances. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria, if the underlying objectives are in conflict rather than in alignment [@ferligoj1992direct]. As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatter plot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The criteria seem to be rather well aligned: There is a high concentration of partitions where both the $\mathit{SSE}$ and the $\mathit{SSE_{Var}}$ have high values. At the same time, there are many partitions that fulfil only one of the criteria very well, but fail to also fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$ by setting $w_t = 1$ ($t = 1, \dots, T$). This simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$ can be dropped. Instead, the tensor is flattened into a two-dimensional matrix by appending the newly computed k-plus variables $\hat{x}_{tij}$ ($t > 1$) to the original data matrix $\mathbf{X}_{N \times P}$. We obtain the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

@@

Thus, when setting the weights $w_t = 1$ ($t = 1, \dots, T$), the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ can be expressed in the same form as the standard k-means criterion $\mathit{SSE}$. Thus, it is possible to define the k-plus criterion by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

Two simulation studies were conducted to evaluate the new k-plus objective for anticlustering. Simulation 1 compared a selected number weighting schemes for the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$. The results of this initial test determined the benchmark implementation of k-plus anticlustering that was used as contestant in the follow-up simulation. Simulation 2 was a large scale comparison of the k-plus objective and other anticlustering objectives striving for between-group similarity: the diversity and the k-means objective.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. In particular, the initial grouping determines the size of the anticlusters, because the exchange procedure following the set-up never changes the sizes of the groups. While the data sets in the simulation studies were always partitioned into equal-sized groups, it would also be possible to ensure that the group sizes differ by 1 at most if an even split is not possible; even different-sized groups can be requested if required by the application.

@@

```{r}
final_tab_sim1 <- round(Reduce("+", results) / length(results) * 100, 2)
rownames(final_tab_sim1) <- c(
  "$\\mathbf{w} = (1, 1, 1, 1)$ / unstandardized",
  "$\\mathbf{w} = (1, 1, 1, 1)$ / standardized",
  "$\\mathbf{w} = (10, 1, 1, 1)$",
  "$\\mathbf{w} = (1, 10, 1, 1)$",
  "$\\mathbf{w} = (1, 1, 10, 1)$",
  "$\\mathbf{w} = (1, 1, 1, 10)$"
)
colnames(final_tab_sim1) <- c(
  "$\\mathit{SSE}$",
  "$\\mathit{SSE_{Var}}$", 
  "$\\mathit{SSE_{Skew}}$",
  "$\\mathit{SSE_{Kurtosis}}$",
  "Average"
)
note <- paste0("Cells indicate how well the maximization of $\\mathit{SSE_{kplusGeneralized}}$ fulfilled each of the four constituting criteria $\\mathit{SSE}$, $\\mathit{SSE_{Var}}$, $\\mathit{SSE_{Skew}}$, and $\\mathit{SSE_{Kurtosis}}$, depending on the weighting scheme. For each simulation run, the relative performance was computed as the $\\mathit{SSE}$ that was obtained by the given weighting scheme, divided by the best $\\mathit{SSE}$ that was obtained by any weighting scheme in this simulation run. Cell values represent the average relative performance across all ", length(results), " simulation runs.")

@@

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. The unweighted standardization approach that performed best in Simulation 1 was employed. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the standard k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.

@@

As a secondary purpose of Simulation 2, I investigated the similarity of partitions obtained when optimizing the different anticlustering objectives. In cluster analysis, partitions identified through different approaches tend to overlap considerably, because similar elements tend to be placed in the same cluster while dissimilar elements tend to be placed in different clusters; this principle holds regardless of the specific clustering method. It is an open question whether the optimization of different anticlustering objectives leads to similar groupings of the elements in anticlusters. To address this question, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each simulation run, i.e., for data set and for each $K$.

@@

[^dispersion]: An exception is given when using the dispersion objective, which ensures that the most similar elements are assigned to different clusters. However, anticlustering objectives striving for overall between-group similarity are mostly unaffected by the grouping of individual item pairs.

>>>>

In the following, I present a variation of the $\mathit{SSE}$ that can be used to minimize differences in variances between groups. To motivate the new criterion, we first consider a set of one-dimensional observations. The sample variance of a set of $N$ observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{x}$:

$$
\mathit{Var(x)} = \frac{1}{N} \sum\limits_{i=1}^{N} (x_i - \overline{x})^2
$$

When defining a new variable $y = (y_1, \ldots, y_N)$ as the squared deviation of each observation from the mean $\overline{x}$, i.e. $y_i = (x_i - \overline{x})^2$, the sample variance of $x$ can be reformulated as the mean of $y$:

@@

By maximizing $\mathit{SSE_{kplus}}$ we seek to obtain a partitioning where means as well as variances are similar between groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of means and variances. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria, if the underlying objectives are in conflict rather than in alignment [@ferligoj1992direct]. As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatter plot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The criteria seem to be rather well aligned: There is a high concentration of partitions where both the $\mathit{SSE}$ and the $\mathit{SSE_{Var}}$ have high values. At the same time, there are many partitions that fulfil only one of the criteria, but fail to simultaneously fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$ by setting $w_t = 1$ ($t = 1, \dots, T$). This simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$ can be dropped. Instead, the tensor is flattened into a two-dimensional matrix by appending the newly computed k-plus variables $\hat{x}_{tij}$ ($t \ge 2$) to the original data matrix $\mathbf{X}_{N \times P}$. We obtain the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

@@

Thus, when setting the weights $w_t = 1$ ($t = 1, \dots, T$), the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ can be expressed in the same form as the standard k-means criterion $\mathit{SSE}$. According to this principle, the k-plus criterion is computed by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and then computing the standard $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

Two simulation studies were conducted to evaluate the new k-plus objective for anticlustering. Simulation 1 compared a selected number of weighting schemes for the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$. The results of this initial test determined the benchmark implementation of k-plus anticlustering that was used as contestant in the follow-up simulation. Simulation 2 was a large scale comparison of the k-plus objective and the most important traditional anticlustering objectives striving for between-group similarity: the diversity and the k-means objective.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. In particular, the initial grouping determines the sizes of the anticlusters, because the exchange procedure following the set-up never changes the sizes of the groups. While the data sets in the simulation studies were always partitioned into equal-sized groups, it would also be possible to ensure that the group sizes differ by 1 at most if an even split is not possible; even different-sized groups can be requested if required by the application.

@@

```{r}
final_tab_sim1 <- round(Reduce("+", results) / length(results) * 100, 2)
rownames(final_tab_sim1) <- c(
  "$\\mathbf{w} = (1, 1, 1, 1)$ / unstandardized",
  "$\\mathbf{w} = (1, 1, 1, 1)$ / standardized",
  "$\\mathbf{w} = (10, 1, 1, 1)$",
  "$\\mathbf{w} = (1, 10, 1, 1)$",
  "$\\mathbf{w} = (1, 1, 10, 1)$",
  "$\\mathbf{w} = (1, 1, 1, 10)$"
)
colnames(final_tab_sim1) <- c(
  "$\\mathit{SSE}$",
  "$\\mathit{SSE_{Var}}$", 
  "$\\mathit{SSE_{Skew}}$",
  "$\\mathit{SSE_{Kurtosis}}$",
  "Average"
)
note <- paste0("Cells indicate how well the maximization of $\\mathit{SSE_{kplusGeneralized}}$ fulfilled each of the four constituting criteria $\\mathit{SSE}$, $\\mathit{SSE_{Var}}$, $\\mathit{SSE_{Skew}}$, and $\\mathit{SSE_{Kurtosis}}$, depending on the weighting scheme. For each simulation run, the relative performance was computed as the result that was obtained by the given weighting scheme, divided by the best result that was obtained by any weighting scheme in this simulation run. Cell values represent the average relative performance across all ", length(results), " simulation runs.")

@@

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. The unweighted standardization approach that performed best in Simulation 1 was employed. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.

@@

As a secondary purpose of Simulation 2, I investigated the similarity of partitions obtained when optimizing the different anticlustering objectives. In cluster analysis, partitions identified through different approaches tend to overlap considerably, because similar elements tend to be placed in the same cluster while dissimilar elements tend to be placed in different clusters; this principle holds regardless of the specific clustering method. However, it is an open question whether the optimization of different anticlustering objectives leads to similar groupings of the elements in anticlusters. To address this question, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each simulation run, i.e., for data set and for each $K$.

@@

[^dispersion]: An exception is given when using the dispersion objective, which ensures that the most similar elements are assigned to different clusters. However, anticlustering objectives striving for overall between-group similarity are largely unaffected by the grouping of individual item pairs.
